---
title: "Practical Machine Learning Course Project"
output: html_document
---

##Overview
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [here.](http://groupware.les.inf.puc-rio.br/har) (see the section on the Weight Lifting Exercise Dataset).


##Admin Issues
The following are the libraries used for this assignment, as well as the seed set, for sake of reproducibility. In particular, the `randomForest` package is more straightforward to use, and `doParallel` allows for higher processing power, and hence lower runtimes.

``` {r library, cache=TRUE, message=FALSE, warning=FALSE}
library(Hmisc)
library(caret)
library(randomForest)
library(foreach)
library(doParallel)
set.seed(123)
```


##Downloading and reading in data

Downloading both datasets from the respective URLs, the test dataset in this case contains the 20 observations to be used for submission later.

``` {r readdata, cache=TRUE, message=FALSE}
train.url    <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
test.url  <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'
download.file(train.url, "train.csv")
download.file(test.url, "test.csv")
```

Removing errorneous and missing entries from the datasets, as well as remove columns which are unused for model building and prediction.

``` {r processdata, cache=TRUE, message=FALSE}
training.df   <-read.csv("train.csv", na.strings=c("NA","#DIV/0!", ""))
test.df <-read.csv("test.csv" , na.strings=c("NA", "#DIV/0!", ""))
training.df<-training.df[,colSums(is.na(training.df)) == 0]
test.df <-test.df[,colSums(is.na(test.df)) == 0]
#Removal of un-used columns
training.df   <-training.df[,-c(1:7)]
test.df <-test.df[,-c(1:7)]
```


##Creating training and test dataset

Splitting the dataset, using 75% for training and 25% for test, as well as identifying the `classe` variable as the outcome variable.

``` {r split, cache=TRUE,message=FALSE}
inTrain    <- createDataPartition(training.df$classe, p = 0.75, list = FALSE)
training.data <- training.df[inTrain, ]
test.data  <- training.df[-inTrain, ]
#Outcome and predictor variables.
y <- training.data$classe
predictors <- training.data[-ncol(training.data)]
```


##Building of model using random forest

Building a random forest model with 800 trees, and using the `doParallel` package here to run on 4 cores.

``` {r model, cache=TRUE,message=FALSE}
registerDoParallel()
model.rf <- foreach(ntree=rep(200, 4), .combine=randomForest::combine, .packages='randomForest') %dopar% {
    randomForest(predictors, y, ntree=ntree)}
```


##Checking accuracy on both training and test datasets

``` {r pred, cache=TRUE, message=FALSE}
train.pred <- predict(model.rf, newdata=training.data)
confusionMatrix(train.pred,training.data$classe)

test.pred <- predict(model.rf, newdata=test.data)
confusionMatrix(test.pred,test.data$classe)
```

For this specific model, the prediction accuracy on the test data set is 99.5%, which is very close to 1. Since the model is performing well, I will proceed to use this on the 20 observations in the evaluation dataset.


##Answer submission script

Following the instructions giving on Coursera, the model is applied on the test dataset to evaluate its performance on 20 new observations. Output is 20 files, each containing the predicted `classe` outcome.

``` {r submission, cache=TRUE, message=FALSE}
pml_write_files = function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

colUsed <- colnames(training.df[colSums(is.na(training.df)) == 0])[-(1:7)]
answer.data <- test.data[colUsed[colUsed!='classe']]
answers<- predict(model.rf,newdata=test.df)

pml_write_files(answers)
```